Photo-D with LSST: stellar photometric distances out to the edge of the Galaxy

### 2024-03-09 paper/methodology.tex
minor edits from Bob

High-precision SDSS photometry clearly shows two white dwarf (WD) sequences in the $g-r$ vs. $u-g$ color-color diagram
(see, e.g., Figures 23 and 24 in \citealt{2007AJ....134..973I}, as well as the middle panel in Figure~\ref{fig:3dataDiags}).
A comparison with models from \cite{1995PASP..107.1047B}, as well as with SDSS spectra, reveals that the two sequences
correspond to H and He white dwarfs, with the mean log(g) = 8.0 for H sequence and log(g) = 8.5 for He sequence.
Upper limits for the scatter of log(g) around these mean sequences appears as no more than 0.5. 
We use three modern white dwarf catalogs: the Montreal White Dwarf
Database\footnote{https://www.montrealwhitedwarfdatabase.org} \citep{2017ASPC..509....3D}, the Gaia EDR3 White Dwarf Catalog\footnote{https://warwick.ac.uk/fac/sci/physics/research/astro/research/catalogues/} \citep{2021MNRAS.508.3877G}, 
and the Gaia-based White Dwarf Database\footnote{https://cdsarc.cds.unistra.fr/viz-bin/cat/J/A+A/679/A127} \citep{2023A&A...679A.127G}
to validate the \cite{1995PASP..107.1047B} models and derive small color offsets that bring models in perfect agreement with these three datasets.

We consider only white dwarfs with cataloged DA, DB or DC spectral class, at galactic latitudes further than 10 degrees from the plane,
with SDSS photometry, apparent magnitudes $5 \le m \le 22$ in any band $m$, and $15 \le r \le 19$, where $r$ is the SDSS $r$-band magnitude.
Observed magnitudes are corrected for interstellar dust extinction using maps from \cite{schlegel_maps_1998} and per-band extinction coefficients
discussed in \S\ref{sec:ismdust} below. We group DB and DC spectral classes as He dwarfs (1,939 objects), while DA spectral class corresponds
to H white dwarfs (9,307 objects). Their absolute magnitudes were calculated using ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}.

@@

We sample the M dwarf luminosity in the range $8.5 \le M_r \le 14.5$, with a step of 0.1 mag, and for each value
generate a track by adding luminosity of a white dwarf, where the track is sampled on the same grid of $M_r$ but
using white dwarf models (separately for H and He models). There are four model families (for two families of white dwarf
models and two values of $[Fe/H]$ for M dwarfs), each with 3,721 $M_r$ vs. color entries. The color tracks for H white dwarfs
and $[Fe/H]=-1.5$ M dwarfs care shown in Figure~\ref{fig:locusBinaries}. Analogous tracks for binaries composed of an M dwarf with
$[Fe/H]=0$ and either an H or He white dwarf look very similar though not identical (differences are at most a few tenths
of a magnitude). In practice, it will be nearly impossible to constrain M dwarf metallicity, and very hard to distinguish
H and He white dwarfs. It may turn out that it will be sufficient to use a single model family to fit LSST data (e.g., a
combination of H white dwarf and $[Fe/H]=-1.5$ M dwarf).

>>>>

High-precision SDSS photometry clearly shows two white dwarf (WD) sequences in the $g-r$ vs. $u-g$ color-color diagram
(see, e.g., Figures 23 and 24 in \citealt{2007AJ....134..973I}, as well as the middle panel in Figure~\ref{fig:3dataDiags}).
A comparison with models from \cite{1995PASP..107.1047B}, as well as with SDSS spectra, reveals that the two sequences
correspond to H and He white dwarfs, with the mean log(g) = 8.0 for the H sequence and log(g) = 8.5 for the He sequence.
Upper limits for the scatter of log(g) around these mean sequences appear as no more than 0.5. 
We use three modern white dwarf catalogs: the Montreal White Dwarf
Database\footnote{https://www.montrealwhitedwarfdatabase.org} \citep{2017ASPC..509....3D}, the Gaia EDR3 White Dwarf Catalog\footnote{https://warwick.ac.uk/fac/sci/physics/research/astro/research/catalogues/} \citep{2021MNRAS.508.3877G}, 
and the Gaia-based White Dwarf Database\footnote{https://cdsarc.cds.unistra.fr/viz-bin/cat/J/A+A/679/A127} \citep{2023A&A...679A.127G}
to validate the \cite{1995PASP..107.1047B} models and derive small color offsets that bring models in perfect agreement with these three datasets.

We consider only white dwarfs with cataloged DA, DB or DC spectral class, at galactic latitudes further than 10 degrees from the plane,
with SDSS photometry, apparent magnitudes $5 \le m \le 22$ in any band $m$, and $15 \le r \le 19$, where $r$ is the SDSS $r$-band magnitude.
Observed magnitudes are corrected for interstellar dust extinction using maps from \cite{schlegel_maps_1998} and per-band extinction coefficients
discussed in \S\ref{sec:ismdust} below. We group the DB and DC spectral classes as He dwarfs (1,939 objects), while the DA spectral class corresponds
to H white dwarfs (9,307 objects). Their absolute magnitudes were calculated using ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}.

@@

We sample the M dwarf luminosity in the range $8.5 \le M_r \le 14.5$, with a step of 0.1 mag, and for each value
generate a track by adding luminosity of a white dwarf, where the track is sampled on the same grid of $M_r$ but
using white dwarf models (separately for H and He models). There are four model families (for two families of white dwarf
models and two values of $[Fe/H]$ for M dwarfs), each with 3,721 $M_r$ vs. color entries. The color tracks for H white dwarfs
and $[Fe/H]=-1.5$ M dwarfs are shown in Figure~\ref{fig:locusBinaries}. Analogous tracks for binaries composed of an M dwarf with
$[Fe/H]=0$ and either an H or He white dwarf look very similar though not identical (differences are at most a few tenths
of a magnitude). In practice, it will be nearly impossible to constrain M dwarf metallicity, and very hard to distinguish
H and He white dwarfs. It may turn out that it will be sufficient to use a single model family to fit LSST data (e.g., a
combination of an H white dwarf and a $[Fe/H]=-1.5$ M dwarf).
